CSE530 Project Final Report\\ Bandwidth-Aware Memory Hierarchy Design with Hybrid Memory Technologies

### 2010-11-18 paper.tex
experiment part added

\documentclass[10pt,letterpaper,singlecolumn]{article}
\usepackage{setspace}
\doublespacing
\usepackage{latex8}
\usepackage{times}
\usepackage{cite}
\usepackage{subfigure}
\usepackage{psfig}
\usepackage[pdftex]{graphicx}
\usepackage{epsfig}
\usepackage{pstricks}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{comment}
\usepackage[small,compact]{titlesec}
\usepackage[small,it]{caption}
\usepackage{wrapfig}

@@

 We will also use
Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator to evaluate our
design method. The benchmarks will be selected from bandwidth-bound
applications, which possibly include benchmarks from SPEC
CPU2006~\cite{SPEC:2006} and NPB~\cite{NPB}.\vspace{0.15in}

@@

@@

\section{Conclusion and Future Work}

In addition, we would like to put an upper bound to the total dynamic energy
consumption of the memory hierarchy. Figure~\ref{fig:memory-energy} shows the
estimation of dynamic power consumptions of different memory technologies with
the provided bandwidths. In figure~\ref{fig:memory-energy}(c), we estimate the total
dynamic energy with 10\% write activities.

>>>>

\documentclass[10pt,letterpaper,singlecolumn]{article}
\usepackage{setspace}
\doublespacing
\usepackage{latex8}
\usepackage{times}
\usepackage{cite}
\usepackage{subfigure}
\usepackage{psfig}
\usepackage[pdftex]{graphicx}
\usepackage{epsfig}
\usepackage{pstricks}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{comment}
\usepackage[small,compact]{titlesec}
\usepackage[small,it]{caption}
\usepackage{wrapfig}
\usepackage{multirow}

@@

Based on the parameters of different cache configurations collected from our
modified version of CACTI~\cite{CACTI}, we evaluate both pure SRAM-based and
hybrid cache hierarchy designs. In this section, we show how bandwidth-aware
memory hierarchy design method help with improving the system performance.

@@

We use Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator in our
experiments. It is configured to model an eight-core CMP. Each core is in-order, and
is similar to UltraSPARC III architecture. The frequency of each core is set to
be 1GHz. Table~\ref{table:cmp-config} lists the detailed parameters of the
baseline. Our baseline contains 8 private L1 instruction and data caches
respectively. Since we only evaluate shared cache hierarchies, each of L1 cache
is fixed to be SRAM-based, and of 64KB capacity. As regard to lower level
caches, we evaluate two cases, pure SRAM-based and hybrid caches with various
memory technologies, including SRAM, MRAM, RRAM, and eDRAM. By evaluating both
cases, we would like to find out optimal cache design that leads to the peak
performance, i.e., the number of cache levels that is reqired, memory technology
used for each level, and capacity of each level.

\begin{table}[t]
  \centering  
  \caption{Baseline CMP configuration.}
  \vspace{0.1in}
  \begin{tabular}{l|c}
    \hline
    \multicolumn{2}{|l|}{\textbf{Core}}\\
    \hline
    No. of cores & 8\\
    \hline
    Frequency & 1GHz\\
    \hline
    Core architecture & in-order, 14-stage pipline, 2 ALUs, 2 FPUs\\
    \hline
    \multicolumn{2}{|l|}{\textbf{Memory}}\\
    \hline    
    \multirow{2}{*}{Private caches} & L1 I-cache: SRAM, 8 x 64KB, 64B line, 2-way, write-through \\
                                    & L1 D-cache: SRAM, 8 x 64KB, 64B line, 2-way, write-through \\
    \hline
    \multirow{2}{*}{Shared caches} & Case 1: Pure SRAM, 64B line, 8-way, write-back \\
                                   & Case 2: Hybrid~(SRAM, MRAM, RRAM, eDRAM), 64B line, 8-way, write-back\\
    \hline    
    Main memory & 4GB\\
    \hline    
  \end{tabular}
  \label{table:cmp-config}
\end{table}

The benchmarks are selected from PARSEC benchmark suite~\cite{PARSEC:2008} with
multithreaded programs, which focus on emerging workloads that are designed to
be representative of next-generation shared-memory programs for CMPs. Since the
performance of different memory technologies are closely related to read and
write intensities, we selected some workloads that vary in the average numbers
of L2 cache read per thousand instructions~(RPKI) and write per thousand
instructions~(WPKI), which are listed in Table~\ref{table:benchmarks}.

\begin{table}[htbp!]
  \centering  
  \caption{Characteristics of selected benchmarks.}
  \vspace{0.1in}
  \begin{tabular}{l|c|c||l|c|c}
    \hline    
    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI} & 
    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI}\\
    \hline
    blackscholes & 22.8 & 61.7 & ferret & 5.7 & 173.4\\
    \hline 
    bodytrack & 5.4 & 139.5 & fluidanimate & 2.6 & 70.1\\
    \hline
    canneal & 5.4 & 29.3 & streamcluster & 17.3 & 16.9\\
    \hline
    facesim & 6.0 & 102.4 & swaptions & 2.6 & 121.4\\
    \hline
  \end{tabular}
  \label{table:benchmarks}
\end{table}

@@

We evaluate various possible configurations with shared caches by simulation,
i.e., with possible numbers of levels, memory technologies, and cache capacities
for each cache level. We consider SRAM and eDRAM as the possible memory
technologies to implement L2 and L3 caches. Both MRAM and RRAM have higher write
latency than SRAM. Furthermore, the endurance of RRAM is too low to be used as
lower level caches. Consequently, these two memory technologies are only
considered to be used as the last level cache.

In the first set of experiments, we evaluate the system performance with
two-level shared caches. Figure~\ref{fig:thput-l2l3} shows the system throughput
with all the benchmarks. It is illustrated that implementing hybrid cache with
eDRAM as the L3 cache helps with performance among most of the benchmarks. Such
performance improvement is more than 10\% in Figure~\ref{fig:thput-l2l3}(a). It
is indicated by our memory technology exploration that eDRAM shows more latency
benefits with larger capacities. As a result, a larger eDRAM-based L3 cache
leads to more performance improvement to the pure SRAM implementation, as
illustrated by Figure~\ref{fig:thput-l2l3}(b). However, hybrid cache does not
always improve the performance, as shown in Figure~\ref{fig:benchmark-l2l3}.
With the benchmark \emph{canneal}, hybrid cache configurations outperform the
pure SRAM implementation among overall range of various capacities. With the
benchmark \emph{ferret}, however, pure SRAM implementation results in higher
performance than hybrid cache implementation with the same capacity
configuration.

\begin{figure*}[t] 
\centering
\includegraphics[width=3in]{figures/thput-l2l3-1}\hspace{0.2in}
\includegraphics[width=3in]{figures/thput-l2l3-2}\\
\hspace{0.02in}
\makebox[3in][l]{\bf (a)}
\makebox[3in][l]{\bf (b)}
\caption{Performance comparison with two-level caches among various benchmarks.
  (a) The L3 cache capacity is 16MB. (b) The L3 cache capacity is 32MB.}
\label{fig:thput-l2l3}
\end{figure*}

\begin{figure*}[t] 
\centering
\includegraphics[width=3in]{figures/canneal-l2l3}\hspace{0.2in}
\includegraphics[width=3in]{figures/ferret-l2l3}\\
\hspace{0.02in}
\makebox[3in][l]{\bf (a)}
\makebox[3in][l]{\bf (b)}
\caption{Performance comparison with two-level caches among various cache
  capacity. (a) The canneal benchmark. (b) The ferret benchmark.}
\label{fig:benchmark-l2l3}
\end{figure*}

In the second set of experiments, we evaluate the system performance with
three-level shared caches. When evaluating hybrid cache configurations, we
consider implementing the last level cache by MRAM and RRAM memory technologies,
since the data transaction intensity is relatively low in the last level cache.
Figure~\ref{fig:thput-l2l3l4} shows the results.
Figure~\ref{fig:thput-l2l3l4}(a) illustrated that the performance more than 15\%
higher on average with a last level cache implemented by MRAM than pure SRAM
implementations. More performance improvement is obtained by increasing the last
level cache capacity, as shown in Figure~\ref{fig:thput-l2l3l4}(b).
Figure~\ref{fig:benchmark-l2l3l4} compares performance with pure SRAM and hybrid
cache implementations with different capacities. In this case, hybrid cache
implementation leads to higher performance in both benchmarks. The results of
the two sets of experiments are illustrated together in
Figure~\ref{fig:benchmark-overall}. An interesting observation is that hybrid
cache implementation shows high performance improvement with one of the
applications \emph{canneal}, whereas as little improvement with the other
\emph{ferret}. The reason is that write latency of both MRAM and RRAM is higher
than SRAM when capacity is less than 1GB. Performance of applications with large
number of writes, such as \emph{ferret} is therefore not benefit from hybrid
cache design.

\begin{figure*}[t] 
\centering
\includegraphics[width=3in]{figures/thput-l2l3l4-1}\hspace{0.2in}
\includegraphics[width=3in]{figures/thput-l2l3l4-2}\\
\hspace{0.02in}
\makebox[3in][l]{\bf (a)}
\makebox[3in][l]{\bf (b)}

\caption{Performance comparison with three-level caches among various
  benchmarks. (a) The L3 cache capacity is 16MB, and the L4 cache capacity is
  512MB. (b) The L3 cache capacity is 32MB, and the L4 cache capacity is 1GB.}
\label{fig:thput-l2l3l4}
\end{figure*}

\begin{figure*}[t] 
\centering
\includegraphics[width=3in]{figures/canneal-l2l3l4}\hspace{0.2in}
\includegraphics[width=3in]{figures/ferret-l2l3l4}\\
\hspace{0.02in}
\makebox[3in][l]{\bf (a)}
\makebox[3in][l]{\bf (b)}
\caption{Performance comparison with three-level caches among various cache
  capacity. (a) The canneal benchmark. (b) The ferret benchmark..}
\label{fig:benchmark-l2l3l4}
\end{figure*}

\begin{figure*}[t] 
\centering
\includegraphics[width=3in]{figures/canneal}\hspace{0.2in}
\includegraphics[width=3in]{figures/ferret}\\
\hspace{0.02in}
\makebox[3in][l]{\bf (a)}
\makebox[3in][l]{\bf (b)}

\caption{Performance comparison with both two-level and three-level caches among
  various cache capacity. (a) The canneal benchmark. (b) The ferret benchmark.}
\label{fig:benchmark-overall}
\end{figure*}

Finally, we evaluate all the possible cache hierarchy configurations with
exhaustive simulations. The optimal cache configurations of each benchmark is
shown in Table~\ref{table:results}. Since the characteristics vary among the
benchmarks, the optimal cache configurations are different among each of
benchmarks. Applications with high write intensities, such as \emph{bodytrack},
\emph{ferret}, and \emph{swaptions} tend to favor SRAM-based caches. Other
applications benefit more from hybrid cache implementation in terms of
performance.

\begin{table}[htbp!]
  \centering  
  \caption{Optimal cache configurations for each benchmark.}
  \vspace{0.1in}
  \begin{tabular}{|l|c|c|c|}
    \hline    
    \textbf{Benchmarks} & \textbf{L2 Configuration} & \textbf{L3 Configuration} & \textbf{L4 Configuration}\\
    \hline
    blackscholes & SRAM, 8MB & eDRAM, 32MB & RRAM, 1GB\\
    \hline 
    bodytrack & SRAM, 8MB & eDRAM, 128MB & -\\
    \hline
    canneal & SRAM, 8MB & eDRAM, 512MB & MRAM, 1GB\\
    \hline
    facesim & SRAM, 8MB & eDRAM, 32MB & MRAM, 1GB\\
    \hline
    ferret & SRAM, 8MB & eDRAM, 32MB & -\\
    \hline
    fluidanimate & SRAM, 8MB & eDRAM, 1GB & -\\
    \hline
    streamcluster & SRAM, 16MB & eDRAM, 128MB & MRAM, 512MB\\
    \hline
    swaptions & SRAM, 8MB & SRAM, 32MB & SRAM, 1GB\\
    \hline
  \end{tabular}
  \label{table:results}
\end{table}

\section{Conclusion and Future Work}
In this project, we explore performance and energy characteristics of various
emerging memory technolgies. Based on our exploration, we evaluate the shared
cache hierarchy design of CMPs optimized for performance by filling the off-chip
memory bandwidth gap. According to our evaluation, SRAM-based caches leads to
reasonable performance with write-intensive applications. Hybrid cache design
help with performance with other types of benchmarks.

Furthermore, a variety of continuing studies is remained to be explored related
to bandwidth-aware memory hierarchy design. First of all, we only examine cache
hierarchy design currently. It is necessary to extend our exploration to the
overall memory hierarchy design. On the other hand, we do not consider energy
efficiency in this project. We would like to put some power constraints to the
memory hierarchy, and examine the energy efficiency of various memory hierarchy
configurations.
